### 2018-01-30 analysis_dtree.tex
preparing for dtree method section

In the section.~\ref{dtree_methodology} we discussed the basics of decision tree method and we provided a brief history of available algorithm. In this section we develop a classification model. Although there are numerous methods to classify labeled data, however, in this study we are interested in a method that not only predict the class of simulation based on metrics, but also give an intuition about the decision process. Therefor we use tree based algorithm to develop the prediction model.\\

Preprocessing data is an important part of data analysis. However, in this study we deal with a clean data without any missing values. All data are reported in the same scale (0-10), therefore, we do not normalize them. After preprocessing data the most important step in developing a prediction model is feature selection. The step becomes more relevant when data set has hundreds to tens of thousands of variable or features. Although our dataset does not have numerous features we do some analysis on features before developing the prediction model. Features of a dataset could have similar behavior. This characteristic is easy to understand by correlation criteria. The Pearson correlation coefficient between two features is defined as

\begin{equation}
R(i)=\frac{cov(X_i,X_j)}{\sqrt{var(X_i)var(X_j)}},
\end{equation}

where $cov$ is covariance function and $var$ is the variance function \citep{Guyon_2003}. Fig.~\ref{fig:Corplot} represent the Pearson Correlation factor among features.

\begin{figure}
    \centering
    \includegraphics
       % [width=\columnwidth]
        [width=200px]
        {figures/pdf/Figure_10.pdf}
    \caption{Pearson correlation values between metrics. The plot is generated using Corrplot package  \citep{Corrplot_2016}  in R programming language.}
    \label{fig:Corplot}
\end{figure}

@@

After analysis of each features the next step is subsetting data for training and testing process. A subset of data that is used for training and test should have about the equal distribution of target values. Equally distributed data helps training process to not biased to an special class also helps to have an accurate evaluation of the model during the test process. Not equally distributed data is called imbalanced. Imbalanced data may decrease the classifier performance. \citet{Branco_2015} provided a comprehensive review of different approaches toward imbalanced distribution of target variables. A mostly common approach is data preprocessing through resampling. According to \citet{Branco_2015}, resampling strategies based on diverse set of techniques such as: random under/over-sampling, distance methods, data cleaning approaches, clustering algorithm among them.\\
According to the clustering results, among all data with different components and velocity model there are 816,1253,879, and 76 data belonging to cluster 1,2,3, and 4, respectively. \citet{Weiss_2003} discussed the fact that the perfectly balanced dataset does not always provide the optimal results. One option could be under sampling cluster 1 to 3 to be in the similar distribution of cluster 4. This may eliminate the useful examples leading to a worse performance. Therefor we 10 times over-sample the data with cluster 4 to have a distribution similar to other cluster. Oversampling may increase the likelihood of overfitting, however, in this case the oversampling rate is not high. On the other hand, using pruning method during training as well as evaluating algorithm with unseen (test) dataset, we address the overfitting issues.  Choosing what fraction of data should be used for training and testing is an open question. We put 30 percent of data for test and 70 percent of data for training process.

@@

In this study we assume $\beta = 1$ to give same importance to precision and recall. \\
Developing a prediction model has a broad discussion. In this study we are not presenting the best model, however, we present an intuition into the method and be able to present a simple model where a user could estimate the target class without using sophisticated algorithms or packages. We divid data into training and test sets. C5.0 package in R has two options to prune the tree. These parameters include: Confidence factor (CF) and the smallest number of samples that must be put in at least two of the splits (minCases). We tune the algorithm for these two parameters based on training with train data and test the results on test data based on F1-score. We also activate the winnow option, which use the internal feature selection algorithm of the C5.0 package. Fig.~\ref{fig:tune} presents the tuning parameters results. It's obvious that the minCases is dominant factor. In this study we choose $CF=0.2$.

\begin{figure}
    \centering
    \includegraphics
       % [width=\columnwidth]
        [width=\textwidth]
        {figures/pdf/Figure_11.pdf}
    \caption{Results of tuning C5.0 algorithm through grid search on confidence factor (CF) and the smallest number of samples that must be put in at least two of the splits (minCases). Color variation represents F1-score. Gray area at C3 has F1-score less than 0.9.}
    \label{fig:tune}
\end{figure}

@@

A decision tree could grow to very low levels and predict all training data accurately unless there is duplicated data with two different target value. The decision tree algorithm in this study provides very good results for the test data. However, the scope of this study is to generate a simple relationship to evaluate the simulation results. In the case of going to very sophisticated methods, other algorithms (e.g., random forest, conditional random forest, neural networks, ...) can be much more accurate. Consequently, In this study we do not activate the boosting capability of the C5.0 algorithm. According to  Fig.~\ref{fig:tuneCFp2} lower minCases and higher CF provides a better results. It is worth to not that the results are based on evaluating algorithm on test data, therefore, higher F1-score are not suggesting overfitting. However, as we mentioned earlier, we do not present the best model, rather we present a simplest model that in general illustrate the feature effects. 
First we start with a very simple model. A simple model needs a strong pruning process.  Therefore we use $minCases = 120, CF=0.2$. Fig.~\ref{fig:dtree1} shows the classification decision tree. 
\begin{figure}
    \centering
    \includegraphics
       % [width=\columnwidth]
        [width=\textwidth]
        {figures/pdf/Figure_13.pdf}
    \caption{First prediction model (M1) for classifying ground motion simulation process based on goodness of fit scores of two metrics (i.e., C8 (Response Spectra) and C4(Total Energy)). Numbers above the box represent the number of training data that end up to that specific node. Bar plots represent the probability of each class under the related condition.  }
    \label{fig:dtree1}
\end{figure}
Table~\ref{tab:confusionmat_test_1} shows  the confusion matrix of applying the prediction model on test dataset and Table~\ref{tab:prec_recall_test_1} presents the precision, recall, and F1-score for each individual class.

@@

\begin{figure}
    \centering
    \includegraphics
       % [width=\columnwidth]
        [width=\textwidth]
        {figures/pdf/Figure_14.pdf}
    \caption{First prediction model (M2) for classifying ground motion simulation process based on goodness of fit scores of two metrics (i.e., C8 (Response Spectra), C4(Total Energy), C6 (Peak Velocity), and C5 (Peak Acceleration)). Numbers above the box represent the number of training data that end up to that specific node. Bar plots represent the probability of each class under the related condition. }
    \label{fig:dtree2}
\end{figure}

>>>>

In the section.~\ref{dtree_methodology} we discussed the basics of decision tree method and we provided a brief history of available algorithm. In this section we develop a classification model. Although there are numerous methods to classify labeled data, however, in this study we are interested in a method that not only predict the class of simulation based on metrics, but also give an intuition about the decision process. Therefor we use tree based algorithm to develop the prediction model.

Preprocessing data is an important part of data analysis. However, in this study we deal with a clean data without any missing values. All data are reported in the same scale (0-10), therefore, we do not normalize them. After preprocessing data the most important step in developing a prediction model is feature selection. The step becomes more relevant when data set has hundreds to tens of thousands of variable or features. Although our dataset does not have numerous features we do some analysis on features before developing the prediction model. Features of a dataset could have similar behavior. This characteristic is easy to understand by correlation criteria. The Pearson correlation coefficient between two features is defined as
% 
\begin{equation}
R(i)=\frac{cov(X_i,X_j)}{\sqrt{var(X_i)var(X_j)}},
\end{equation} 
% 
where $cov$ is covariance function and $var$ is the variance function \citep{Guyon_2003}. Fig.~\ref{fig:Corplot} represent the Pearson Correlation factor among features.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/pdf/Figure_10.pdf}
    \caption{Pearson correlation values between metrics. The plot is generated using Corrplot package  \citep{Corrplot_2016}  in R programming language.}
    \label{fig:Corplot}
\end{figure}

@@

After analysis of each features the next step is subsetting data for training and testing process. A subset of data that is used for training and test should have about the equal distribution of target values. Equally distributed data helps training process to not biased to an special class also helps to have an accurate evaluation of the model during the test process. Not equally distributed data is called imbalanced. Imbalanced data may decrease the classifier performance. \citet{Branco_2015} provided a comprehensive review of different approaches toward imbalanced distribution of target variables. A mostly common approach is data preprocessing through resampling. According to \citet{Branco_2015}, resampling strategies based on diverse set of techniques such as: random under/over-sampling, distance methods, data cleaning approaches, clustering algorithm among them.

According to the clustering results, among all data with different components and velocity model there are 816,1253,879, and 76 data belonging to cluster 1,2,3, and 4, respectively. \citet{Weiss_2003} discussed the fact that the perfectly balanced dataset does not always provide the optimal results. One option could be under sampling cluster 1 to 3 to be in the similar distribution of cluster 4. This may eliminate the useful examples leading to a worse performance. Therefor we 10 times over-sample the data with cluster 4 to have a distribution similar to other cluster. Oversampling may increase the likelihood of overfitting, however, in this case the oversampling rate is not high. On the other hand, using pruning method during training as well as evaluating algorithm with unseen (test) dataset, we address the overfitting issues.  Choosing what fraction of data should be used for training and testing is an open question. We put 30 percent of data for test and 70 percent of data for training process.

@@

In this study we assume $\beta = 1$ to give same importance to precision and recall.

Developing a prediction model has a broad discussion. In this study we are not presenting the best model, however, we present an intuition into the method and be able to present a simple model where a user could estimate the target class without using sophisticated algorithms or packages. We divid data into training and test sets. C5.0 package in R has two options to prune the tree. These parameters include: Confidence factor (CF) and the smallest number of samples that must be put in at least two of the splits (minCases). We tune the algorithm for these two parameters based on training with train data and test the results on test data based on F1-score. We also activate the winnow option, which use the internal feature selection algorithm of the C5.0 package. Fig.~\ref{fig:tune} presents the tuning parameters results. It's obvious that the minCases is dominant factor. In this study we choose $CF=0.2$.

\begin{figure*}
    \centering
    \includegraphics
       % [width=\columnwidth]
        [width=\textwidth]
        {figures/pdf/Figure_11.pdf}
    \caption{Results of tuning C5.0 algorithm through grid search on confidence factor (CF) and the smallest number of samples that must be put in at least two of the splits (minCases). Color variation represents F1-score. Gray area at C3 has F1-score less than 0.9.}
    \label{fig:tune}
\end{figure*}

@@

A decision tree could grow to very low levels and predict all training data accurately unless there is duplicated data with two different target value. The decision tree algorithm in this study provides very good results for the test data. However, the scope of this study is to generate a simple relationship to evaluate the simulation results. In the case of going to very sophisticated methods, other algorithms (e.g., random forest, conditional random forest, neural networks, ...) can be much more accurate. Consequently, In this study we do not activate the boosting capability of the C5.0 algorithm. According to  Fig.~\ref{fig:tuneCFp2} lower minCases and higher CF provides a better results. It is worth to not that the results are based on evaluating algorithm on test data, therefore, higher F1-score are not suggesting overfitting. However, as we mentioned earlier, we do not present the best model, rather we present a simplest model that in general illustrate the feature effects.

First we start with a very simple model. A simple model needs a strong pruning process.  Therefore we use $minCases = 120, CF=0.2$. Fig.~\ref{fig:dtree1} shows the classification decision tree.

\begin{figure*}
    \centering
    \includegraphics
       % [width=\columnwidth]
        [width=\textwidth]
        {figures/pdf/Figure_13.pdf}
    \caption{First prediction model (M1) for classifying ground motion simulation process based on goodness of fit scores of two metrics (i.e., C8 (Response Spectra) and C4(Total Energy)). Numbers above the box represent the number of training data that end up to that specific node. Bar plots represent the probability of each class under the related condition.  }
    \label{fig:dtree1}
\end{figure*}
Table~\ref{tab:confusionmat_test_1} shows  the confusion matrix of applying the prediction model on test dataset and Table~\ref{tab:prec_recall_test_1} presents the precision, recall, and F1-score for each individual class.

@@

\begin{figure*}
    \centering
    \includegraphics
       % [width=\columnwidth]
        [width=\textwidth]
        {figures/pdf/Figure_14.pdf}
    \caption{First prediction model (M2) for classifying ground motion simulation process based on goodness of fit scores of two metrics (i.e., C8 (Response Spectra), C4(Total Energy), C6 (Peak Velocity), and C5 (Peak Acceleration)). Numbers above the box represent the number of training data that end up to that specific node. Bar plots represent the probability of each class under the related condition. }
    \label{fig:dtree2}
\end{figure*}

### 2018-01-30 decision-tree.tex
clustring first pass revised
reduced clustering section
preparing for dtree method section

\subsection{Decision Tree Construction}
\label{sec:decision-tree}

Working...

% ===========================================================================================
%
% OLD NAEEM VERSION
% 
% Decision tree learning is a method for approximating a target function, in which learned function is represented by decision tree \citep{Mitchell_1997_Book}. \citet{Quinlan_1986} demonstrated that the technology for building decision trees from examples is fairly robust. He summarized an approach to synthesizing decision trees that has been used in a variety of systems, and he describes one such system, ID3, in detail. Although ID3 algorithm is successful in most classification problems, it does not handle the numeric attributes and only one attribute at a time is tested for making decisions. \citet{Quinlan_1993} extended the ID3 algorithm and presented the C4.5 algorithm. Later on they added boosting capability to C4.5 and called it C5.0. The improved algorithm accepts both continues and discrete features and solves over fitting problem by pruning. Although there are many other algorithm for implementing decision tree, in this study we use the C5.0 algorithm. Discussing the algorithm in details is out of the scope of this paper. There are fairly good amount of references to study those details \citep[e.g.,][]{Mitchell_1997_Book,Quinlan_1993,Hornik_2009}.\\
% In general, decision tree measures the effectiveness of an attribute in classifying the training data through information gain. According to  \citet{Mitchell_1997_Book}, information gain of an attribute is the expected reduction in entropy caused by partitioning the examples according to this attribute. If the target can take on $c$ different values the entropy is defined as
% \begin{equation}
% Entropy(S) = \sum_{i=1}^{c} -p_{i}\ log_2\ p_i,
% \end{equation}

% where $p_i$ is proportion of $S$ belonging to class $i$. Entropy measures the homogeneity of examples or dataset. Higher entropy means data is  fairly uniformly distributed among different classes. On the other hand lower entropy means data unequally distributed among classes. The extreme case happens when all data belong to one class which results in entropy equal to zero. Given entropy as a measure of impurity in a collection of training example the measure of the effectiveness of attribute is defined through information gain which is expected reduction in entropy caused by partitioning the examples according to this attribute. Gain of the attribute A is defined as

% \begin{equation}
% Gain(S,A)\ =\ Entropy(S) - \sum_{v \in Values(A) } \frac{|S_v|}{|S|} Entropy(S_v)
% \end{equation}

% where $Values(A)$ is the set of all possible values of attribute A, and $S_v$ is the subset of $S$ for which attribute $A$ has value $v$. Information gain is used by the algorithm in order to determine which is the better attribute for classifying the training example. The attribute with higher information gain will be the first option to separate the training data based on that attribute and grow to the lower nodes (Refer to chapter 3 of \citet{Mitchell_1997_Book} for more detailed examples.) \\
% We discussed the basic idea behind the decision tree algorithm, however, there are more details specially in dealing with outliers and noises which we address them through pruning approaches. In this study we use C5.0 package \citep{C50_2015} in R programming language \citep{R_2016_program} to develop the decision tree algorithms. We discuss the input parameters in the section.\ref{classification_model}.

>>>>

\subsection{Decision Tree Design}
\label{sec:decision-tree}

The construction of a decision tree is a supervised learning process to obtain a target function used to measure the effectiveness of an attribute to classify data. There are a good number of approaches to develop decision trees \citep[e.g.,][]{Quinlan_1986_ML, Mitchell_1997_Book, Hornik_2009_}

We use the C4.5 algorithm developed by \citet{Quinlan_1993_Book}...

% In machine learning, decision trees are classified as a supervised method,

% ===========================================================================================
%
% OLD NAEEM VERSION
% 
Decision tree learning is a method for approximating a target function, in which learned function is represented by decision tree \citep{Mitchell_1997_Book}. \citet{Quinlan_1986} demonstrated that the technology for building decision trees from examples is fairly robust. He summarized an approach to synthesizing decision trees that has been used in a variety of systems, and he describes one such system, ID3, in detail. Although ID3 algorithm is successful in most classification problems, it does not handle the numeric attributes and only one attribute at a time is tested for making decisions. \citet{Quinlan_1993_Book} extended the ID3 algorithm and presented the C4.5 algorithm. Later on they added boosting capability to C4.5 and called it C5.0. The improved algorithm accepts both continues and discrete features and solves over fitting problem by pruning. Although there are many other algorithm for implementing decision tree, in this study we use the C5.0 algorithm. Discussing the algorithm in details is out of the scope of this paper. There are fairly good amount of references to study those details \citep[e.g.,][]{Mitchell_1997_Book, Quinlan_1993_Book, Hornik_2009_}.

In general, decision tree measures the effectiveness of an attribute in classifying the training data through information gain. According to  \citet{Mitchell_1997_Book}, information gain of an attribute is the expected reduction in entropy caused by partitioning the examples according to this attribute. If the target can take on $c$ different values the entropy is defined as
% 
\begin{equation}
Entropy(S) = \sum_{i=1}^{c} -p_{i}\ log_2\ p_i,
\end{equation}
% 
where $p_i$ is proportion of $S$ belonging to class $i$. Entropy measures the homogeneity of examples or dataset. Higher entropy means data is  fairly uniformly distributed among different classes. On the other hand lower entropy means data unequally distributed among classes. The extreme case happens when all data belong to one class which results in entropy equal to zero. Given entropy as a measure of impurity in a collection of training example the measure of the effectiveness of attribute is defined through information gain which is expected reduction in entropy caused by partitioning the examples according to this attribute. Gain of the attribute A is defined as 
% 
\begin{equation}
Gain(S,A)\ =\ Entropy(S) - \sum_{v \in Values(A) } \frac{|S_v|}{|S|} Entropy(S_v)
\end{equation} 
% 
where $Values(A)$ is the set of all possible values of attribute A, and $S_v$ is the subset of $S$ for which attribute $A$ has value $v$. Information gain is used by the algorithm in order to determine which is the better attribute for classifying the training example. The attribute with higher information gain will be the first option to separate the training data based on that attribute and grow to the lower nodes (Refer to chapter 3 of \citet{Mitchell_1997_Book} for more detailed examples.)

We discussed the basic idea behind the decision tree algorithm, however, there are more details specially in dealing with outliers and noises which we address them through pruning approaches. In this study we use C5.0 package \citep{C50_2015} in R programming language \citep{R_2016_program} to develop the decision tree algorithms. We discuss the input parameters in the section \ref{classification_model}.
